---
title: "qPCR analysis template"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

**.Rmd script** 

Script workflow:  
1. Limit of detection (LOD) and limit of quantification (LOQ) calculations based on a standard curve  
2. Reading in datafiles and metadata  
3. Evaluating No Template Controls (NTCs; Negatives)  
4. 
 

## Load libraries

```{r}
library(tidyverse)   # covers ggplot2, dplyr, tidyr, readr, purrr, stringr
library(readxl)      # Excel input
library(writexl)     # Excel output
library(ggrepel)     # text repulsion in plots
library(drc)         # LOD modeling
library(ggpubr)      # publication-quality plots
library(strex)       # extra string utilities if you need them
```

## Setting paths prior to code 

```{r}
### Data 
std_curve_path="example standard curve.csv"
raw_data_path="example input/qPCR_results/"
metadata_path="example input/client_metadata_example.xlsx"

### Plots
std_curve_plot="example output/Standard_curve.png"
inhibition_plot="example output/Inhibition.png"
blanks_piechart="example output/Blanks_piechart.png"
field_samples_piechart="example output/Fields_piechart.png"
field_samples_barchart="example output/Fieldsamples_barchart.png"
outliers_plot="example output/Outliers.png"

### Results
results_blank="example output/Results_Blanks.xlsx"
results_samples="example output/Results.xlsx"
```

## Standard Curve Information 

Read in data and confirm No Template Controls (NTC)s contain no data 

```{r}
std_curve_raw <- read.csv(std_curve_path, skip = 19) 

std_curve_raw %>% filter(Sample == "Neg")
```

Calculate mean Cq and remove No Template Controls (NTC)s 

```{r}
std_curve <- std_curve_raw %>% dplyr::rename(St_Quant=7) %>%
  
  ## Remove NTCs 
  filter(!Sample == "Neg") %>%
  
  ## Calculate mean Cq
  dplyr::group_by(Sample) %>%
  mutate(Cq_mean = mean(Cq, na.rm=TRUE))
```

Calculating standard curve metrics

```{r}
# Fit a linear model to calculate slope and intercept
linear_model <- lm(Cq_mean ~ log10(St_Quant), data = std_curve)

# Extract slope and y-intercept
slope <- coef(linear_model)[2]       # Slope (m)
y_intercept <- coef(linear_model)[1] # Y-intercept (b)
efficiency <- (-1+(10^(-1/slope)))

## Standard curve calculations (assuming std_curve is already grouped by Sample)
std_curve <- std_curve %>%
  
  ## calculate detection rates (number of Cqs present / number of replicates)
  mutate(detection_rate = sum(!is.na(Cq)) / n(),
  
  ## calculate Cq stats - SD, and CV
         Cq_sd = sd(Cq, na.rm=TRUE),
         #Cq_cv = (Cq_sd / Cq_mean) * 100 ## old method 
         Cq_cv = sqrt(((1+efficiency) ^ ((Cq_sd^2) * log(1+efficiency))) - 1)) %>%
  ungroup()
```

### Calculating LOD and LOQ based on discrete thresholds

```{r}
# Assuming std_curve is your data frame
quants <- std_curve %>% dplyr::select(St_Quant) %>% distinct() %>% arrange(desc(St_Quant))

# Creating a factored list of these standard curve values
quant_list <- as.factor(quants$St_Quant)
```

Find the Limit of Detection (LOD) based on 95% detection rate 

```{r}
# Find the first value below the 0.95 detection rate
first_below_LOD_threshold <- std_curve %>% filter(detection_rate < 0.95) %>% arrange(desc(St_Quant)) %>%
  slice(1) %>% dplyr::select(St_Quant)

# Get the standard quantity of the first value below the threshold
below_threshold_LOD_quant <- first_below_LOD_threshold$St_Quant[1]

# Find the index of this quantity in the ordered list
LOD_index <- which(quants$St_Quant == below_threshold_LOD_quant)

LOD <- quants$St_Quant[LOD_index - 1]
```

Find the Limit of Quantification (LOQ) based on CV 35% 

```{r}
# Find the first value below the 0.35 CV
first_above_LOQ_threshold <- std_curve %>% filter(Cq_cv > 0.350) %>% arrange(desc(St_Quant)) %>%
  slice(1) %>% dplyr::select(St_Quant)

# Get the standard quantity of the first value below the threshold
above_threshold_LOQ_quant <- first_above_LOQ_threshold$St_Quant[1]

# Find the index of this quantity in the ordered list
LOQ_index <- which(quants$St_Quant == above_threshold_LOQ_quant)

LOQ <- quants$St_Quant[LOQ_index - 1]
```

### Using modeling to calculate LOD and LOQ

We're troubleshooting this, ignore for now 

```{r}
## Different models 
# 
# LOQ1 <- nls(Cq.CV~SSasymp(log10(Standards),Asym,R0,lrc),
#                   data=DAT2[DAT2$Target==Targets[i],])
# 
# 
# LOQ2 <- lm(Cq.CV~log10(Standards),data=DAT2[DAT2$Target==Targets[i],])
# 
# 
# LOQ3 <- lm(Cq.CV~poly(log10(Standards),2),data=DAT2[DAT2$Target==Targets[i],])
# 
# LOQ4 <- lm(Cq.CV~poly(log10(Standards),3),data=DAT2[DAT2$Target==Targets[i],])
# 
# LOQ5 <- lm(Cq.CV~poly(log10(Standards),4),data=DAT2[DAT2$Target==Targets[i],])
# 
# LOQ6 <- lm(Cq.CV~poly(log10(Standards),5),data=DAT2[DAT2$Target==Targets[i],])
# 
# LOQ7 <- lm(Cq.CV~poly(log10(Standards),6),data=DAT2[DAT2$Target==Targets[i],])

## find residual standard error
## pick best fitting model 
```

### Plotting standard curve 

```{r}
std_curve %>% 
  ## Taking only those detections with >50% to make curve 
  filter(detection_rate > 0.50) %>%
  
  ## Plotting that std. curve 
  ggplot(., aes(x = log10(St_Quant), y = Cq)) +
  
  geom_smooth(method = 'lm', se = FALSE, color = "darkred") +

  stat_regline_equation(size = 4, color = 'darkred',
                        aes(label = after_stat(rr.label)), hjust=1,
                        label.x.npc = "right", label.y.npc = "top") +
  
  geom_point(color = 'black', fill = 'white', shape = 21, size = 3.5, alpha=0.8) +
  
  ## ADDING LOD LINE -- if no line shows up, all st. curve is qualitative 
  geom_vline(xintercept=log10(LOD), linetype="dashed", color = "grey40") +
  geom_text(aes(x = log10(LOD), y = max(std_curve$Cq, na.rm=TRUE) * 0.3,
                label = paste("LOD =", LOD, "copies")), 
            angle = 90, vjust = -0.5, hjust = -0.1, color = "grey40", size = 4) +
  
  ## ADDING LOQ LINE -- if no line shows up, all st. curve is quantitative 
  geom_vline(xintercept=log10(LOQ), linetype="dashed", color = "grey40") +
  geom_text(aes(x = log10(LOQ), y = max(std_curve$Cq, na.rm=TRUE) * 0.3, 
                label = paste("LOQ =", LOQ, "copies")), 
            angle = 90, vjust = -0.5, hjust = -0.1, color = "grey40", size = 4) +

  theme_bw() +
  
  labs(
    x = "Normalized standard concentrations",
    y = "Cq"
  ) +
  
  theme(axis.text.y = element_text(size=10, color="grey20"),
        axis.text.x = element_text(size=10, color="grey20"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=12, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=12, face="bold"))

ggsave(std_curve_plot, width=6, height=4)
```

## Reading in datafiles and merging into one large dataframe

```{r}
### USER EDITS:
## In str_after_nth(file.ID, "results/", 1), make sure this matches the folder ending referenced in list.files

df <- 
  # list files in directory following a particular pattern
  list.files(path = raw_data_path, pattern = ".csv", full.names = TRUE) %>%
  
  # get the column names
  set_names(.) %>% 
  
  # join all files together in one data frame by file ID, skipping first 19 rows
  map_dfr(~read.csv(., skip = 19), .id = "file.ID") %>% 
  
  # turn file.ID into just plate information (plate.ID)
  mutate(file.ID = str_after_nth(file.ID, "results/", 1),
         file.ID = str_before_nth(file.ID, ".csv", 1)) %>%
  dplyr::rename(plate.ID = file.ID, Sample_ID = Sample) %>%
  
  #### HOLD SPACE FOR CUSTOM SAMPLE ID NEEDS, 
  ##### IF NONE, COMMENT THIS SECTION OUT 
  ## RI STRIPED BASS SPECIFIC CODE FOR SAMPLE ID MISMATCHES
  mutate(Sample_ID = gsub(" #1", "_1", Sample_ID),
         Sample_ID = gsub(" #2", "_2", Sample_ID)) %>%
  
  mutate(Sample_ID = case_when(
    str_detect(Well, "4") & Sample_ID == "2/21/24_ONE_NTR" ~ paste0(Sample_ID, "_1"),
    str_detect(Well, "5") & Sample_ID == "2/21/24_ONE_NTR" ~ paste0(Sample_ID, "_2"),
    TRUE ~ Sample_ID
  )
  ##### #########################
  
  )

head(df)
```

## Reading in meta df

Sample ID from meta needs to match the Sample ID from the qPCR output

```{r}
meta <- read_xlsx(metadata_path) %>%
  
  #### HOLD SPACE FOR CUSTOM SAMPLE ID NEEDS, 
  ##### IF NONE, COMMENT THIS SECTION OUT 
  ## RI STRIPED BASS SPECIFIC CODE FOR SAMPLE ID ISSUES
  ## For other projects, address any Sample_ID differences (Sample_ID on qPCR output needs to match meta)
  mutate(Sample_ID = gsub(" #1", "_1", Sample_ID),
         Sample_ID = gsub(" #2", "_2", Sample_ID))
```

## Evaluating No Template Controls (NTCs; Negatives)

```{r}
NTC <- df %>% 
  filter(grepl("neg", Sample_ID, ignore.case = TRUE))

## Calculate number of negatives with Cq values 
sum(!is.na(NTC$Cq))

## Calculate number of negatives total 
nrow(NTC)

## Calculate number of plates
length(unique(NTC$plate.ID))
```

## Spike information

For each sample, calculate the distance from the Positive Spike-in value.

```{r}
spike_samples <- df %>% 
  ## subset to spiked samples 
  filter(grepl("spike|pos", Sample_ID, ignore.case = TRUE)) %>%
  
  ## remove spike from SampleID column only if 'pos' is NOT in the Sample_ID
  mutate(Sample_ID = case_when(
    !grepl("pos", Sample_ID, ignore.case = TRUE) ~ str_before_nth(Sample_ID, " spike", 1),
    TRUE ~ Sample_ID
  )) %>%

  ## group by plate ID and sample ID 
  group_by(plate.ID, Sample_ID) %>%

  ## calculate mean of spikes per plate and sample 
  ## ungroup by Sample_ID so the next calculation is only done grouped by Plate 
  mutate(Filter_Cq = mean(Cq)) %>% 
  ungroup(Sample_ID) %>%

  ## calculate difference from plate's Cq value 
  mutate(Cq_diff = Filter_Cq - Filter_Cq[grepl("pos", Sample_ID, ignore.case = TRUE)]) %>%
  ungroup(); spike_samples

#### HOLD SPACE FOR CUSTOM SAMPLE ID NEEDS, 
##### IF NONE, COMMENT THIS SECTION OUT
## Remove any lab error samples 
## RI Striped Bass only 
## removing 2/21/24 DI Blank 1 (pipetting issue) 
spike_samples <- spike_samples %>%
  filter(!Sample_ID == "2/21/24_Tap Blank_1")
```

Plot those distance values and save to output folder

```{r}
# Create a jitter position object
jitter_pos <- position_jitter(width = 0.45, seed = 123)

## Plot samples 
spike_samples %>% dplyr::select(Target, Sample_ID, Cq_diff) %>% distinct() %>%
  
  ggplot(., aes(x=Target, y = Cq_diff)) + 
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", color = "grey50", size=1.5) +
  
  geom_jitter(aes(fill = abs(Cq_diff) > 2), 
              size = 2, alpha = 0.5, color = 'black', shape = 21, #width = 0.45,
              position = jitter_pos) +
  
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "white")) +
  geom_text_repel(
    aes(label = ifelse(abs(Cq_diff) > 2, Sample_ID, "")),  position = jitter_pos,
    size = 3, box.padding = 0.5, point.padding = 0.2, force = 2
  ) +
  
  labs(
    x = "Sample",
    y = "Distance from Positive Control"
  ) +
  
  theme_bw() +
  
  ## keep y axis at least -2,2 but extend to max 
  coord_cartesian(
    ylim = c(
      min(-2.5, min(spike_samples$Cq_diff, na.rm = TRUE)),
      max(2.5, max(spike_samples$Cq_diff, na.rm = TRUE))
    )) +
  
  ## theme variables
    theme(panel.background=element_rect(fill='white', colour='black'),
        legend.position = "none",
        axis.text.y = element_text(size=10, color="grey20"),
        axis.text.x = element_blank(),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=12, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=12, face="bold"))

ggsave(inhibition_plot, width=4, height=4)
```

## Filter data Cq, Number of Replicates, and Copy Num calculations

```{r}
## Complete calculations
filters_df <- df %>% 
  ## subset out the spiked samples 
  filter(!grepl("spike|pos|neg", Sample_ID, ignore.case = TRUE)) %>%

  ## group by plate ID and sample ID 
  group_by(plate.ID, Sample_ID) %>%

  ## calculate mean of Cq per plate and sample, replacing NaN with NA
  ## calculate # of replicates 
  ## calculate copy number 
  reframe(Filter_Cq = if_else(is.nan(mean(Cq, na.rm = TRUE)), NA_real_, mean(Cq, na.rm = TRUE)),
         Filter_Num_Replicates = sum(!is.na(Cq)),
         Filter_Copy_Num = 10^((Filter_Cq-y_intercept)/slope)
         ) 
```

## Addressing duplicate samples (a.k.a., re-runs)

This code assumes the user wants to always use the most recent plate if completing a sample for a 2nd time.  
*If this is not the case, chat with Fisheries team to make sure code reflect user’s needs.*

```{r}
filters_df <- filters_df %>%
  ## Separating Plate.ID into Date and Plate Number
  separate(plate.ID, c("Plate_date", "Plate_number"), sep = " PLATE") %>%
  
  ## Change Date column to a Date format 
  mutate(Plate_date = dmy(Plate_date)) %>%
  
  ## Group by Sample_ID and keep only the row with the most recent date
  group_by(Sample_ID) %>%
  slice_max(Plate_date, n = 1, with_ties = FALSE) %>% ungroup()

## confirm the number of rows matches the number of unique Sample IDs (output should be TRUE)
nrow(filters_df) == length(unique(filters_df$Sample_ID))
```

## Combing with meta and collapsing by sample

Sometimes samples have 2 filters and the data needs to be the mean of those two filters for each water sample. 

```{r}
samples_df <- filters_df %>% right_join(meta, ., by = "Sample_ID") %>%
  group_by(Date, Sample_Location) %>%
  
  ## mutate to mean Ct value
  mutate(`Mean Ct` = if_else(is.nan(mean(Filter_Cq, na.rm = TRUE)), 
                                    NA_real_, mean(Filter_Cq, na.rm = TRUE)),
         
         ## take highest value of number of replicates
         `Number of Replicates` = max(Filter_Num_Replicates, na.rm=TRUE),
         
         ## sum of 2 copy number values 
         `Mean Copy Number` = ifelse(sum(Filter_Copy_Num, na.rm=TRUE) == 0, NA, 
                            sum(Filter_Copy_Num, na.rm=TRUE))
           ) %>%
  
  ungroup() %>%
  
  ## summarizing df 
  dplyr::select(Date, Sample_Location, Sample_Type, Number_of_Filters, 
                `Mean Ct`, `Number of Replicates`, `Mean Copy Number`) %>%
  distinct() %>%

  ## adding new SampleID back in
  unite(Sample_ID, Date, Sample_Location, sep = " ", remove=F)

head(samples_df)

## Confirm that the number of samples output from qPCR matches the number of samples in the metadata (output = TRUE)
nrow(samples_df) == nrow(meta %>% dplyr::select(-Sample_ID) %>% distinct())
```

## Normalizing data

```{r}
## normalize data 
normalized_df <- samples_df %>%
  ## take log10 of copy number 
  mutate(`Mean Copy Number Normalized` = log10(`Mean Copy Number` + 1))
        
### Identifying outliers
outliers <- normalized_df %>% dplyr::select(`Sample_ID`, `Mean Copy Number Normalized`) %>%
  mutate(
    Median = median(`Mean Copy Number Normalized`),
    MAD = mad(`Mean Copy Number Normalized`, constant=1),

    Mi = 0.6745*(abs(`Mean Copy Number Normalized` - Median)/MAD),
    Outlier = ifelse(Mi > 3.5, "Outlier", ""),


    Upper_threshold = (3.5*MAD+0.6745*Median)/0.6745,
    Lower_threshold =-1*(3.5*MAD-0.6745*Median)/0.6745
  ) %>%

  arrange(desc(`Mean Copy Number Normalized`))

unique(outliers$Upper_threshold)

## create an outlier cut-off 
cutoff <- quantile(normalized_df$`Mean Copy Number Normalized`, na.rm = TRUE, probs=0.75) + 
  1.5*IQR(normalized_df$`Mean Copy Number Normalized`, na.rm=TRUE)

## create an outlier cut-off 
cutoff_below <- quantile(normalized_df$`Mean Copy Number Normalized`, na.rm = TRUE, probs=0.25) - 
  1.5*IQR(normalized_df$`Mean Copy Number Normalized`, na.rm=TRUE)

## Output the values that outside the cutoff
normalized_df %>% filter(`Mean Copy Number Normalized` > cutoff) 
normalized_df %>% filter(`Mean Copy Number Normalized` < cutoff_below) 

## the cutoff we move forward with is the cutoff (high) but confirm no values are below the lower cutoff value 
```

## Adding present/absent information

```{r}
normalized_df <- normalized_df %>%
  mutate(Detection = case_when(
    is.na(`Mean Copy Number Normalized`) ~ "Absent",
    TRUE ~ "Present"
  ))
```

## Blank specific information

```{r}
blank_df <- normalized_df %>% subset(Sample_Type == "Blank")

detection_counts <- blank_df %>%
  count(Detection) %>%
  mutate(percentage = n / sum(n) * 100,
         label = paste0(Detection, "\n", "n=", n, "\n", round(percentage, 1), "%"))

ggplot(detection_counts, aes(x = "", y = n, fill = Detection)) +
  geom_bar(stat = "identity", width = 1, color = "black", alpha=0.75) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = label), 
            position = position_stack(vjust = 0.5), 
            size = 5,
            fontface = "bold") + 
  theme_void() +
  theme(legend.position = "none",
        #plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
        ) +
  scale_fill_manual(values = c("#e9ecef", "#669bbc"))

ggsave(blanks_piechart, width=4, height=4)
```

## Sample preliminary data

Pie chart

```{r}
fieldsamples_df <- normalized_df %>% subset(Sample_Type == "Field")

field_detection_counts <- fieldsamples_df %>%
  count(Detection) %>%
  mutate(percentage = n / sum(n) * 100,
         label = paste0(Detection, "\n", "n=", n, "\n", round(percentage, 1), "%"))

ggplot(field_detection_counts, aes(x = "", y = n, fill = Detection)) +
  geom_bar(stat = "identity", width = 1, color = "black", alpha=0.75) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = label), 
            position = position_stack(vjust = 0.5), 
            size = 5,
            fontface = "bold") + 
  theme_void() +
  theme(legend.position = "none",
        #plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
        ) +
  scale_fill_manual(values = c("#e9ecef", "#669bbc"))

ggsave(field_samples_piechart, width=4, height=4)
```

Bar chart

```{r}
fieldsamples_df %>% subset(Detection == "Present") %>%
  
  ggplot(., aes(x=Sample_ID, y=`Mean Copy Number Normalized`)) + 
  geom_bar(stat = "identity", width = 0.7, alpha=0.5, fill = "#669bbc") +
  labs(
    y="Log10-Normalized Copy Number",
    x = "Sample ID",
  ) +
  
  ## Outlier line
  geom_hline(yintercept = cutoff, color = "darkred", linetype = "dashed", size = 0.5) +
  geom_text(aes(y = cutoff, x = 2, label = "Outlier"), hjust = 0.4, vjust=-0.5, color = "grey40", size = 3) +
  
  ## LOD line
  geom_hline(yintercept=log10(LOD), linetype="dashed", color = "grey40") +
  geom_text(aes(y = log10(LOD), x = 2, label = "LOD"), hjust = 0.5, vjust=-0.5, color = "grey40", size = 3) +
  
  ## LOQ line
  geom_hline(yintercept=log10(LOQ), linetype="dashed", color = "grey40") +
  geom_text(aes(y = log10(LOQ), x=2, label = "LOQ"), hjust = 0.5, vjust=-0.5, color = "grey40", size = 3) +
  
  theme_bw() +
    ## theme variables
    theme(panel.background=element_rect(fill='white', colour='black'),
        #legend.position = c(0.98, 0.98),  # This puts the legend in the top right corner
        legend.position = "right",
        legend.justification = c(1, 1),  # This aligns the legend to the top right
        axis.text.y = element_text(size=8, color="grey20"),
        axis.text.x = element_text(size=6, color="grey20", angle=90, vjust=0.5, hjust=1),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=10, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold"),
        # New theme elements for strip appearance
        strip.background = element_rect(fill = "white", color = "black"),
        strip.text = element_text(face = "bold", size = 9)
    )

ggsave(field_samples_barchart, width = 9.5, height=5)
```

Show distance of outliers

```{r}
fieldsamples_df %>% ggplot(., aes(x=Sample_Type, y=`Mean Copy Number`, 
                                  fill = ifelse(is.na(`Mean Copy Number Normalized`) | is.na(cutoff), FALSE, 
                                          `Mean Copy Number Normalized` > cutoff))) + 
  geom_jitter(width=0.2, size=3.5, color='black', shape=21, alpha=0.5) +
  scale_fill_manual(values = c("#545E56", "#c1121f"),
                    labels = c("Normal", "Outlier")) +
  theme_bw() +
  #geom_hline(yintercept = cutoff, linetype = "dotted", color = "grey50") +
  labs(x = "Sample",
       fill = "Outlier Detection"
       ) +
      theme(panel.background=element_rect(fill='white', colour='black'),
        legend.position = "right",
        axis.text.y = element_text(size=8, color="grey20"),
        axis.text.x = element_blank(),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=10, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold"),
    ) +
    guides(fill = guide_legend(override.aes = list(alpha = 1)))

ggsave(outliers_plot, width = 5, height=4.5)
```

## Exporting data

```{r}
blank_df %>% mutate(Date = as.Date(Date)) %>% 
  dplyr::select(Date, Sample_Location, `Mean Ct`, `Number of Replicates`, 
                `Mean Copy Number`, `Mean Copy Number Normalized`, Detection) %>%
  
  unite(Sample, Date, Sample_Location, sep = " ", remove = TRUE) %>%
  
  # cut decimals down to 2
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  
  write_xlsx(results_blank)

#### HOLD SPACE FOR CUSTOM SAMPLE ID NEEDS, 
##### DELETE IF NOT NEEDED 
### adding some meta for Rich 
rich_meta <- read_xlsx("C:/BoxDrive/Box/Science/Fisheries/Projects/eDNA/RI Striped Bass/metadata/eDNA_Data_RI_STB_2024.xlsx") %>%
  dplyr::rename(Date = Sample_Date, Sample_Location = Station_Code) %>%
  mutate(Sample_Time = format(Sample_Time, format = "%H:%M:%S"))

normalized_df %>% #full_join(rich_meta, ., by = c("Date", "Sample_Location")) %>%
  mutate(Date = as.Date(Date)) %>%
  dplyr::select(-Number_of_Filters) %>% write_xlsx(results_samples)
```





