---
title: "Metabarcoding data quality: eDNA metabarcoding base script"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

This script evaluates your sequence quality and taxonomic assignment quality. Figures produced in this script can go into supplemental data for a manuscript. 

# Load libraries

```{r}
library(dplyr) # for data transformation
library(tidyverse) # for data transformation
library(ggplot2) # for plotting
library(readxl) ## for reading in excel files
library(viridis)
library(hrbrthemes)
library(ggrepel)
library(cowplot)

# removing scientific notation
## remove line or comment out if not desired 
options(scipen=999)
```

# Load data 

```{r}
### User edits:
### 1. Replace the 3 paths below: edit example_input to your project specific path 
### 2. Confirm your sampleIDs match between metadata, results df, and filtering stats output

filtering_stats <- read_tsv("example_input/overall_summary.tsv", show_col_types = FALSE) %>% dplyr::rename(SampleID = sample)

meta <- read.csv("example_input/metadata.csv", header=TRUE)

results <- read_xlsx("example_output/Results_rawreads_long.xlsx") %>%
  mutate(Category = factor(Category, levels = c("Human", "Livestock", "Other", "unassigned", "Bird",
                                                "Elasmobranch", "Marine Mammal", "Sea Turtle", "Teleost Fish")))
                                                       
ASV_breakdown <- read_xlsx("example_output/ASV_breakdown.xlsx") %>%
  mutate(Category = factor(Category, levels = c("Human", "Livestock", "Other", "unassigned", "Bird",
                                                "Elasmobranch", "Marine Mammal", "Sea Turtle", "Teleost Fish")))
```

# Sequence data

## Data Transformation 

No user edits. 

```{r}
df <- full_join(filtering_stats, meta, by = "SampleID") %>%
  # filtering out columns we don't need 
  dplyr::select(-cutadapt_reverse_complemented) %>%
  
  # removing percentage icon from cutadapt_passing_filters_percent
  mutate(cutadapt_passing_filters_percent = gsub("%", "", cutadapt_passing_filters_percent)) %>%
  
  # confirming that all columns of interest are numerical 
  mutate_at(vars(2:10), as.numeric) %>%
  
  # data transformation so all columns of interest are together 
  gather("measure", "value", 2:10)  
```

## Plotting 

Suggested webpage to choose colors: https://coolors.co/

```{r}
### User edits:
### 1. Change paths of output to desired folder (data/figures is suggested data structure)
### 2. Change x axis and color, size based on metadata desired 
### 3. Change custom colors and sizes if desired and number of colors and sizes based on metadata variable chosen 

df %>% 
  ## USER EDITS IN LINE BELOW 
  ggplot(., aes(x=Project, y=value, color=Site, shape=SampleType)) + 
  
  ## adding points in jitter format 
  geom_jitter(width=0.15, alpha=0.5) + 
  
  ## option for additional boxplots if desired (uncomment to add)
  #geom_boxplot() +
  
  ## using facet_wrap to create grid based on variables and factor() to order them in custom format
  facet_wrap(~factor(measure, levels=c('cutadapt_total_processed', 'cutadapt_passing_filters', 
                                       'cutadapt_passing_filters_percent', 'DADA2_input',
                                 'filtered', 'denoisedF', 'denoisedR', 'merged', 'nonchim')), scales = "free") +
  
  ## graph asthetics 
  theme_bw() +
  ylab("Number of reads") + 
  
  ## USER EDITS IN MANUAL CODE BELOW 
  scale_color_manual(values = c("red3", "lightblue", "purple2", "gold", "green4", "black")) +
  scale_size_manual(values = c(21,17)) +
  
  
  theme(panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='black'),
        strip.text = element_text(size = 10, face="bold"),
        legend.position = "right",
        axis.text.y = element_text(size=7, color="grey30"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=11, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=11, face="bold"))

ggsave("example_output/Figures/SampleReport_FilteringStats.png", width = 10, height=8)
```


# Plot unassigned taxonomy

## Data transformation 

No user edits.

```{r}
results_summary <- results %>% 
  group_by(Category, Project) %>%
  summarise(sum_reads = sum(reads))

general_stats <- results %>% 
  group_by(Category) %>%
  summarise(sum_reads = sum(reads)) %>% ungroup() %>%
  mutate(total = sum(sum_reads),
         percent = sum_reads/total*100) %>% dplyr::select(Category, percent) %>% distinct() %>%
  ## round to 2 decimal places 
  mutate(across(c('percent'), round, 4))

ASV_summary <- ASV_breakdown %>%
  group_by(Category) %>%
  summarise(count = n_distinct(ASV_ID))

species_summary <- results %>%
  group_by(Category) %>%
  summarise(count = n_distinct(Species_name))

## species in addition to category
## metadata option add-in
```


## Raw Reads Plotting 

With metadata 

```{r}
### User edits:
### 1. Change paths of output to desired folder (data/figures is suggested data structure)
### 2. Change x axis and color, size based on metadata desired 
### 3. Change custom colors and sizes if desired and number of colors and sizes based on metadata variable chosen 
### 4. Comment out any categories that do not show up in your dataset

# Check how many categories 
unique(results_summary$Category)
## Based on this output, comment/uncomment the categories present for color 

ggplot(results_summary, aes(fill=Category, y=sum_reads, x=Project)) + 
    geom_bar(position="stack", stat="identity") +
    scale_fill_brewer(palette = "RdYlBu") +
    # scale_fill_manual(values = c("#9f040e", # Human
    #                              "#e30613", # Livestock
    #                              "#fb747d", # Other
    #                              "#ff0000", # unassigned
    #                              "#03045e", # Bird
    #                              "#023e8a", # Elasmobranch
    #                              "#0077b6", # Marine mammal
    #                              "#0096c7", # Sea Turtle
    #                              "#48cae4" # Teleost Fish 
    #                              )) +
    labs(fill = "Category") +
    theme_bw() + 
    xlab(" Month") + ylab("Raw reads") +
    theme(panel.background=element_rect(fill='white', colour='black'),
        axis.text.y = element_text(size=7, color="grey30"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=11, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=11, face="bold"))

## add % stacked as well

ggsave("example_output/Figures/Rawreads_percategory.png", width = 8, height = 5)
```


Reads Piechart 

```{r}
### User edits:
### 1. Change paths of output to desired folder (data/figures is suggested data structure)
### 2. Change scale brewer color if desired 

piechart <- general_stats %>%  
  mutate(csum = rev(cumsum(rev(percent))), 
         pos = percent/2 + lead(csum, 1),
         pos = if_else(is.na(pos), percent/2, pos))

piechart_reads <- general_stats %>% 
  ggplot(., aes(x="", y = percent, fill = Category)) +
  geom_col(color = "black") +
  geom_label_repel(data = piechart,
                   aes(y = pos, label = paste0(percent, "%")),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "RdYlBu") +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    axis.title = element_blank(),     # Remove axis titles
    legend.position = "none"
  ) +
  ggtitle("% of raw reads") +
  xlab("") + ylab("") + labs(fill = "Category")
```


ASV Piechart 

```{r}
### User edits:
### 1. Change paths of output to desired folder (data/figures is suggested data structure)
### 2. Change scale brewer color if desired 

piechart_ASV <- ASV_summary %>%  
  mutate(csum = rev(cumsum(rev(count))), 
         pos = count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), count/2, pos))

piechart_ASV <- ASV_summary %>% 
  ggplot(., aes(x="", y = count, fill = Category)) +
  geom_col(color = "black") +
  geom_label_repel(data = piechart_ASV,
                   aes(y = pos, label = paste0(count)),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "RdYlBu") +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    axis.title = element_blank(),     # Remove axis titles
    legend.position = "none"
  ) +
  ggtitle("# of ASVs") +
  xlab("") + ylab("") + labs(fill = "Category")
```

Number of species pie chart 

```{r}
piechart_spp <- species_summary %>%  
  mutate(csum = rev(cumsum(rev(count))), 
         pos = count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), count/2, pos))

piechart_spp <- species_summary %>% 
  ggplot(., aes(x="", y = count, fill = Category)) +
  geom_col(color = "black") +
  geom_label_repel(data = piechart_spp,
                   aes(y = pos, label = paste0(count)),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "RdYlBu") +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    axis.title = element_blank()     # Remove axis titles
  ) +
  ggtitle("# of species") +
  xlab("") + ylab("") + labs(fill = "Category")
```

Plot together and export 

```{r}
plot_grid(piechart_reads, piechart_ASV, piechart_spp, 
          ncol=3, 
          rel_widths = c(2,2,3.075) 
          #align = "hv"
          )

ggsave("example_output/Category_breakdown.png", width=10, height=4)
```



















