---
title: "Relative Abundance Heatmaps: eDNA metabarcoding base script"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

**.Rmd script** 

This script plots your relative abundance matrix as a heatmap. Figures produced are potentially part of the main figures of your manuscript/report. 

## Load libraries

```{r}
library(ggplot2) ## for plotting
library(dplyr) ## for data table manipulation
library(tidyr) ## for data table manipulation
library(readxl) ## for reading in excel files
library(stringr) ## for data transformation
library(strex) ## for data transformation
library(purrr) ## for data transformation
library(funrar) ## for make_relative()
library(tidyverse) ## for data transformation
library(naniar) ## replace_with_na_all function
library(ggh4x) ## for facet wrap options
library(tidytext)
library(forcats)
library(scales)
```

## Load data 

```{r}
df <- read_xlsx("example_output/Results_relab_long.xlsx") %>%
  mutate(across(c(relab), ~ round(.x, 5)))

taxlevels <- read_excel(
  "C:/BoxDrive/Box/Science/Fisheries/Projects/eDNA/Metabarcoding Lab Resources/Reference Databases/GMGI_Vert_Ref.xlsx") %>% 
  dplyr::select("Species_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "species") %>%
  distinct()

df_annotated <- df %>% left_join(., taxlevels, by = "Species_name")

## bringing in common names information for those not in our db
commonNames_annotated <- read_xlsx("example_output/Taxonomic_assignments/CommonNames_required_edited.xlsx")
  
# Loop through each row of the dataframe to add taxonomic level information from required edited worksheet 
for (i in commonNames_annotated$Species_name) {
  # Extract the current row (will do this for each ASV_ID in the choice df)
  current_row <- commonNames_annotated %>% subset(Species_name==i)
  
  # Apply filter based on the current row's condition
df_annotated <- df_annotated %>%
  mutate(across(c(Common_name, Category, Kingdom, Phylum, Class, Order, Family, Genus, species),
                ~case_when(Species_name == current_row$Species_name ~ current_row[[cur_column()]],
                           TRUE ~ .x)))
}
```

## Remove Categories 

If you want to plot relative abundance without human, other, or livestock categories. As FYI/warning, relative abundance is calculated with these categories included. Relative abundance can also be thought of as proportion of total reads, which is calculated from the total reads for that sample. 

```{r}
df_filtered <- df_annotated %>% 
  filter(!Category == "Other" & !Category == "Livestock" & !Category == "Unassigned" & !Category == "Human") 

df_average <- df_annotated %>%
  group_by(SampleType, Species_name) %>%
  ### average relative abundance by sample Type
  mutate(avg_relab = mean(relab, na.rm=TRUE))
```

If targeted species heatmap is desired, replace df_annotated with df_filtered in the heatmap code below. If so, remember to change the file name exported. 

## Plot 

reverse label order: scale y discrete limits reverse limits=rev

https://coolors.co/ (hit tools on the top right hand side)

```{r}
## if subset of categories is desired, replace df below with df_filtered
df_annotated %>%
  
  ## replace zeros with NAs for plotting
  replace_with_na_all(condition = ~.x == 0.00000) %>%
  
  # Create a factor for Common_name ordered by Order within each Category
  group_by(Category) %>%
  mutate(Common_name = factor(Common_name, levels = unique(Common_name[order(Order, desc(Common_name))]))) %>%
  ungroup() %>%
  
  ## ggplot basic options (USER EDIT: X AND Y AXIS)
  ggplot(., aes(x = sampleID, y = Common_name)) +
  geom_tile(aes(fill = relab), color = "black") +
  
  ## x, y, and legend labels (USER EDITS IF DESIRED)
  ylab("Common name") +
  xlab("Site") +
  labs(fill = "Relative Abundance (%)") +
  
  ## color of the tile options; direction=1 will flip the low/high (USER EDITS IF DESIRED)
  scale_fill_gradient(na.value = "white", low = "lightskyblue2", high = "#0C4D66") + 
  
  ## facet grid with Category and project variables
  facet_grid2(Category ~ SampleType, 
              scales = "free", space = "free", 
              labeller = labeller(Category = label_wrap_gen(width = 10))) +
  
  ## graph theme options
  theme_classic() +
  theme(
    ## axis text 
    axis.text.x = element_text(angle = 90, size=6, color="grey25", hjust = 1),
    axis.text.y = element_text(colour = 'black', size = 8),
    
    ## legend text and title 
    legend.text = element_text(size = 8, color="black"),
    legend.title = element_text(margin = margin(t = 0, r = 0, b = 5, l = 0), size=10, color="black", face="bold"),
    legend.position = c(-0.4, -0.05), 
    legend.key.height = unit(5, 'mm'),
    legend.direction = "horizontal",
    legend.key.width = unit(5, 'mm'),
    legend.title.align = 0.5,
    legend.title.position = "top",
    
    ## axis titles 
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=14, face="bold"),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=14, face="bold"),
        
    ## facet wrap labels
    strip.text.x = element_text(color = "black", face = "bold", size = 12),
    strip.text.y = element_text(color = "black", face = "bold", size = 12, angle=0),
    strip.background.y = element_blank(),
    strip.clip = "off"
    )
   
  
## USER EDITS WIDTH AND HEIGHT TO DESIRED   
ggsave("example_output/Figures/Relative_abundance.png", width = 22, height = 14)  
```

Heatmap plot by sample type 

```{r}
df_average %>%
  dplyr::select(Species_name, Common_name, Category, SampleType, Kingdom:avg_relab) %>%
  
  ## replace zeros with NAs for plotting
  replace_with_na_all(condition = ~.x == 0.00000) %>%
  
  # Create a factor for Common_name ordered by Order within each Category
  group_by(Category) %>%
  mutate(Common_name = factor(Common_name, levels = unique(Common_name[order(Order, desc(Common_name))]))) %>%
  ungroup() %>%
  
    ## ggplot basic options (USER EDIT: X AND Y AXIS)
  ggplot(., aes(x = SampleType, y = Common_name)) +
  geom_tile(aes(fill = avg_relab), color = "black") +
  
  ## x, y, and legend labels (USER EDITS IF DESIRED)
  ylab("Common name") +
  xlab("") +
  labs(fill = "Relative Abundance (%)") +
  
  ## color of the tile options; direction=1 will flip the low/high (USER EDITS IF DESIRED)
  scale_fill_gradient(na.value = "white", low = "lightskyblue2", high = "#0C4D66") + 
  
  ## facet grid with Category and project variables
  facet_grid2(Category ~ SampleType, 
              scales = "free", space = "free", 
              labeller = labeller(Category = label_wrap_gen(width = 10))) +
  
  ## graph theme options
  theme_classic() +
  theme(
    ## axis text 
    axis.text.x = element_text(angle = 90, size=6, color="grey25", hjust = 1),
    axis.text.y = element_text(colour = 'black', size = 8),
    
    ## legend text and title 
    legend.text = element_text(size = 8, color="black"),
    legend.title = element_text(margin = margin(t = 0, r = 0, b = 5, l = 0), size=10, color="black", face="bold"),
    legend.position = c(-0.4, -0.05), 
    legend.key.height = unit(5, 'mm'),
    legend.direction = "horizontal",
    legend.key.width = unit(5, 'mm'),
    legend.title.align = 0.5,
    legend.title.position = "top",
    
    ## axis titles 
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=14, face="bold"),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=14, face="bold"),
        
    ## facet wrap labels
    strip.text.x = element_text(color = "black", face = "bold", size = 12),
    strip.text.y = element_text(color = "black", face = "bold", size = 12, angle=0),
    strip.background.y = element_blank(),
    strip.clip = "off"
    )

ggsave("example_output/Figures/Relative_abundance_sampletype.png", width = 7, height = 14) 
```

## Top Species List visual 

```{r}
top_list <- read_xlsx("example_output/Results_rawreads_long.xlsx") %>%
  filter(!Category == "Other" & !Category == "Livestock" & !Category == "unassigned" & !Category == "Human") %>%
  group_by(Species_name) %>%
  summarise(total = sum(reads),
            log = log10(total)) %>% 
  arrange(desc(total)) %>%
  head(25) 

ggplot(top_list, aes(x = fct_reorder(Species_name, log), y = log)) +
  geom_segment(aes(xend = Species_name, yend = 0), color = "#006e90") +  # Lollipop stick
  geom_point(size = 3, color = "#006e90") +  # Lollipop head
  coord_flip() +  # Flip coordinates for horizontal lollipop chart
  labs(
       x = "",
       y = "Normalized Reads") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8, face="italic"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=12, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=12, face="bold")) +
  scale_y_continuous(
    labels = comma,
    limits = c(0, max(top_list$log) + 1.5)  # Set the upper limit to max value + 2
  )

ggsave("example_output/Figures/Top_species.png", width=6, height=6)
```






