---
title: "Sub-sampling and Decontamination"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

**.Rmd script** 

This script provides code for optional sub-sampling and decontamination based on NTCs and blanks. 

## Load libraries

```{r}
library(tidyverse) # for data transformation
library(ggplot2) # for plotting
library(readxl) ## for reading in excel files
library(viridis)
library(hrbrthemes)
library(ggrepel)
library(cowplot)
library(Rmisc)
library(writexl)
library(vegan) ## rareify 
library(patchwork)
library(microDecon)

# removing scientific notation
## remove line or comment out if not desired 
options(scipen=999)
set.seed(123)
```

## Load data 

For the sake of demonstration, I added a NTC to the results matrix. Change the file name for own data.

```{r}
### loading RData if applicable uncomment
# load("../results/Raw_data.RData")
# data <- ASV_table_taxID_collapsed

data <- read_xlsx("example_output/Results_initial/Results_rawreads_matrix_mockNTC.xlsx") %>%
  dplyr::rename()

meta <- read.csv("example_input/SBNMS_metadata.csv") %>%
  mutate(sampleID = gsub("-", "_", sampleID))
```

## Read Counts prior to decontamination

Visualize the read count distribution. There is one sample that has more reads than the rest of the samples (>200,000 in this case).

```{r}
## Calculate total read numbers 
totals <- data.frame(
  sampleID = colnames(data)[4:ncol(data)],   # column names for sample counts
  nonchim  = colSums(data[, 4:ncol(data)])   # total reads per sample
) %>% left_join(., meta, by = "sampleID")

total_presub <- totals %>% ggplot(., aes(x=SampleType, y=nonchim)) +
  geom_boxplot(outlier.shape = NA, aes(color=SampleType)) +
  labs(
    x = "Sample Type", y = "Total reads pre sub-sample"
  ) +
  geom_jitter(aes(fill=SampleType), shape=21, size=1.5, alpha=0.5, width=0.25) + 
  theme_bw() +
    theme(
      legend.position = "none"
    ); total_presub
```

## Sub-sample above 200,000

Change this cut-off to what you'd like it to be based on your data

```{r}
## Compute total reads per sample
sample_depths <- colSums(data[, 4:ncol(data)])  # each column is a sample
summary(sample_depths)

## Define target depth
### CUSTOMIZE VALUE AS NEEDED 
target_depth <- quantile(sample_depths, 0.99)
```

Edit based on the target depth above 

```{r}
spp_info <- data[, 1:3]
counts <- data[, 4:ncol(data)]
counts_t <- t(counts)

counts_adj <- counts_t
high_depth <- rowSums(counts_t) > target_depth

n_high <- sum(high_depth)

if (n_high == 1) {
  counts_adj[high_depth, ] <- rrarefy(
    matrix(counts_t[high_depth, ], nrow = 1),
    sample = target_depth
  )
} else if (n_high > 1) {
  counts_adj[high_depth, ] <- rrarefy(counts_t[high_depth, ], sample = target_depth)
}

counts_adj <- t(counts_adj)
df_adj <- cbind(spp_info, counts_adj)
```

Confirming that only the desired samples were changed 

```{r}
new_totals <- data.frame(
  sampleID = colnames(df_adj)[4:ncol(df_adj)],   # column names for sample counts
  new_nonchim  = colSums(df_adj[, 4:ncol(df_adj)])   # total reads per sample
)

totals_adj <- totals %>% full_join(new_totals) %>%
  mutate(difference = new_nonchim-nonchim)

## confirm no species were lost (output should be empty)
zero_rows <- which(rowSums(df_adj[, 4:ncol(df_adj)]) == 0)
```

## Visualizing afterwards

```{r}
total_after_sub <- totals_adj %>% ggplot(., aes(x=SampleType, y=new_nonchim)) +
  geom_boxplot(outlier.shape = NA, aes(color=SampleType)) +
  labs(
    x = "Sample Type", y = "Total reads post sub-sample"
  ) +
  geom_jitter(aes(fill=SampleType), shape=21, size=1.5, alpha=0.5, width=0.25) + 
  theme_bw() +
    theme(
      legend.position = "none"
    ); total_after_sub

combined_plot <- total_presub + total_after_sub + plot_layout(ncol = 2)
ggsave("example_output/Figures/Filtering_subsample_before_after.png", combined_plot, width = 10, height = 5, dpi = 300)
```


### Microdecon 

https://github.com/donaldtmcknight/microDecon

At this point, remove positive controls from the dataset (not applicable to this example).

```{r}
## Add annotation as needed for the name of the NTC or blank
df_for_microDecon <- df_adj %>%
  dplyr::select(-(Common_name:Category)) %>%
  relocate(matches("NTC|neg|Blank"), .after = Species_name)

## negative and blanks, columns 2
## sample columns 3:144

decontaminated <- decon(data = df_for_microDecon, 
                        
                        ## number of blank columns [integer]
                        numb.blanks=1, 
                        
                        ## number of individuals per group [integer]
                        ## number of samples - 1 (still troubleshooting why this is - 1..)
                        numb.ind=c(141),
                        
                        ## Indicate if taxa columns are present (T/F)
                        taxa=T)

removed <- decontaminated$reads.removed
save(removed, file = "example_output/Results_decontaminated/Decontaminated_reads_removed.RData")

decontamin_df <- decontaminated$decon.table

## if multiple blanks, a new column called Mean.blank is created, if not the column name remains the same
decontamin_df <- decontamin_df %>% 
  #dplyr::select(-Mean.blank)
  dplyr::select(-NTC)

df_filtered <- df_adj %>% dplyr::select(Species_name:Category) %>% right_join(., decontamin_df)

nrow(df_adj) ## Number spp before microdecon
nrow(df_filtered) ## Number spp after microdecon
```

Exporting dataframe 

```{r}
df_filtered %>% write_xlsx("example_output/Results_decontaminated/Results_read_count_decontaminated.xlsx")

df_relab <- df_filtered %>% 
  gather("sampleID", "reads", 4:last_col()) %>%
  dplyr::group_by(sampleID) %>%

  ### total
  dplyr::mutate(sample_total = sum(reads)) %>%
  dplyr::group_by(sampleID, Species_name) %>%

  ## relab
  dplyr::mutate(relab = reads/sample_total) %>% ungroup() %>%
  select(-reads, -sample_total) 

df_relab %>% write_xlsx("example_output/Results_decontaminated/Results_relative_abundance_decontaminated.xlsx")
```


