---
title: "Summarizing data for report"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

**.Rmd script** 

## Load libraries

```{r}
library(ggplot2) ## for plotting
library(tidyverse) ## for data table manipulation
library(readxl) ## read in excel spreadsheet 
library(ggh4x) ## for facet wrap options
library(writexl)
library(ggrepel)
library(cowplot)
library(scales)

library(phyloseq)     # ecological stats
library(microbiome)   # alpha diversity
```

## Load data 

Change the paths for your exported excel matrices. 

```{r}
species_breakdown <- read_xlsx("docs/eDNA 12S metab/example_output/Summary_Species_level.xlsx")
asv_breakdown <- read_xlsx("docs/eDNA 12S metab/example_output/Summary_ASV_level.xlsx")

raw_reads <- read_xlsx("docs/eDNA 12S metab/example_output/Results_matrix.xlsx") %>% 
  gather("sampleID", "reads", 4:last_col())

rel_ab_df <- read_xlsx("docs/eDNA 12S metab/example_output/Results_matrix_relative.xlsx") %>% 
  gather("sampleID", "reads", 4:last_col())

## bringing in common names information for those not in our db
commonNames_annotated <- read_xlsx("docs/eDNA 12S metab/example_output/taxonomic_assignments/CommonNames_required_edited.xlsx")
```

Read in gmgi db taxonomic level information and additional common names from this project.   
No user edits.   
  
```{r}
taxlevels <- read_excel(
  "C:/BoxDrive/Box/Science/Fisheries/Projects/eDNA/Metabarcoding Lab Resources/Reference Databases/GMGI_Vert_Ref.xlsx") %>% 
  dplyr::select("Species_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "species", "PhylogenySorting") %>%
  distinct()

df_annotated <- raw_reads %>% left_join(., taxlevels, by = "Species_name")

# Loop through each row of the dataframe to add taxonomic level information from required edited worksheet 
for (i in commonNames_annotated$Species_name) {
  # Extract the current row (will do this for each ASV_ID in the choice df)
  current_row <- commonNames_annotated %>% subset(Species_name==i)
  
  # Apply filter based on the current row's condition
df_annotated <- df_annotated %>%
  mutate(across(c(Common_name, Category, Kingdom, Phylum, Class, Order, Family, Genus, species, PhylogenySorting),
                ~case_when(Species_name == current_row$Species_name ~ current_row[[cur_column()]],
                           TRUE ~ .x)))
}

df_tax <- df_annotated %>% dplyr::select(Species_name, Kingdom:PhylogenySorting) %>% distinct()
```

## Defining categories and order 

```{r}
## Print the unique entries in the column Category
unique(rel_ab_df$Category) 

### order these how you'd like the report to be ordered (the figures)
categories = c(
  "Unassigned", "Human", "Livestock", "Mammal", 
  "Bird", "Reptile",
  "Marine Mammal", "Teleost Fish"
  )

## re-ordering all dfs based on the category list above
species_breakdown <- species_breakdown %>% dplyr::mutate(Category = factor(Category, levels = categories))
asv_breakdown <- asv_breakdown %>% dplyr::mutate(Category = factor(Category, levels = categories))
rel_ab_df <- rel_ab_df %>% dplyr::mutate(Category = factor(Category, levels = categories))
raw_reads <- raw_reads %>% dplyr::mutate(Category = factor(Category, levels = categories))
```

Confirm that you didn't forget a category. If you did, <NA> will appear in the list. Re-run the code above *AND* the read_xlsx() code chunk to fix if needed. 

```{r}
unique(rel_ab_df$Category) 
```

## Graph color and order options

Edit colors and/or comment/uncomment as needed based on your category list. 

```{r}
categories

fill_colors <- c(
  "Unassigned" = "grey85",
  "Human" = "#FE9E20", 
  "Livestock" = "#FCCA46",
  "Mammal" = "#FFECC2",
  "Bird" = "#C8D2B1",
  "Reptile" = "#A5B57D",
  "Marine Mammal" ="#DCF1F9",
  #"Elasmobranch" = "#97ADCB",
  "Teleost Fish" = "#85B6CB"
  )
```

## Calculating data for pie charts 

No user edits. 

```{r}
### Creating a results summary df
## Group by category and calculate the total reads per category
results_summary <- raw_reads %>% 
  dplyr::group_by(Category) %>%
  reframe(sum_reads = sum(reads))

### Creating general stats df 
## calculate the percent of total reads per category / total reads of sequencing run
general_stats <- results_summary %>% 
  dplyr::mutate(total = sum(sum_reads),
         percent = sum_reads/total*100) %>% 
  dplyr::select(Category, percent) %>% distinct() %>%
  ## round to 2 decimal places 
  dplyr::mutate(across(c('percent'), round, 4))

### Creating ASV Summary df that will go into ASV pie chart 
## Group by category and count the # of distinct(unique) ASVs per category
ASV_summary <- asv_breakdown %>%
  dplyr::group_by(Category) %>%
  reframe(count = n_distinct(ASV_ID))

### Creating species summary df that will go into species pie chart
## Group by category and calculate the number of distinct(unique) tax IDs per category
species_summary <- raw_reads %>%
  dplyr::group_by(Category) %>%
  reframe(count = n_distinct(Species_name))
```

### Generate % reads piechart 

User edits:    
1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
### Create pie chart df that calculates 'pos' and 'csum'
## 'csum' = reverse cumulative sum of percent (sum of that row’s percent plus all rows after it)
## 'pos' = Computes a label position. this ensures the pie chart is ordered by % total 
piechart <- general_stats %>%
  dplyr::mutate(csum = rev(cumsum(rev(percent))), 
         pos = percent/2 + lead(csum, 1),
         pos = if_else(is.na(pos), percent/2, pos)) %>%
  dplyr::mutate_if(is.numeric, round, digits = 3)

## create pie chart 
## legend is removed on this plot for the report generation; user can add it back in if desired 
piechart_reads <- general_stats %>% 
  ggplot(., aes(x="", y = percent, fill = Category)) +
  geom_col(color = "black", width=1.25) +
  geom_label_repel(data = piechart,
                   aes(y = pos, label = paste0(percent, "%")),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = fill_colors) +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
    legend.position = "none",
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    axis.title = element_blank()     # Remove axis titles
      ) +
  ggtitle("Raw reads (%)") +
  xlab("") + ylab("") + labs(fill = "Category"); piechart_reads

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Category_breakdown_percent_rawreads.png", width=4, height=3)
```

### Generate ASV pie chart 

This is for internal use, not the contract report. 

User edits:    
1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
piechart_ASV <- ASV_summary %>%  
  dplyr::mutate(csum = rev(cumsum(rev(count))), 
         pos = count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), count/2, pos))

ASV_summary %>% 
  ggplot(., aes(x="", y = count, fill = Category)) +
  geom_col(color = "black", width=1.25) +
  geom_label_repel(data = piechart_ASV,
                   aes(y = pos, label = paste0(count)),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = fill_colors) +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    #legend.position = "none"
    axis.title = element_blank()     # Remove axis titles
      ) +
  ggtitle("Number of ASVs") +
  xlab("") + ylab("") + labs(fill = "Category")

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Category_breakdown_ASVs.png", width=4, height=3)
```

### Generate the number of species pie chart

User edits:    
1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
piechart_spp <- species_summary %>%  
  dplyr::mutate(csum = rev(cumsum(rev(count))), 
         pos = count/2 + lead(csum, 1),
         pos = if_else(is.na(pos), count/2, pos))

piechart_species_plot <- species_summary %>% 
  ggplot(., aes(x="", y = count, fill = Category)) +
  geom_col(color = "black", width=1.25) +
  geom_label_repel(data = piechart_spp,
                   aes(y = pos, label = paste0(count)),
                   size = 3, nudge_x = 1, show.legend = FALSE) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = fill_colors) +
  theme_bw() +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
    panel.border = element_blank(),  # Remove panel border
    panel.grid = element_blank(),    # Remove grid lines
    axis.ticks = element_blank(),    # Remove axis ticks
    axis.text = element_blank(),     # Remove axis text
    axis.title = element_blank()     # Remove axis titles
  ) +
  ggtitle("Number of Taxonomic Assignments") +
  xlab("") + ylab("") + labs(fill = "Category"); piechart_species_plot

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Category_breakdown_Species.png", width=4, height=3)
```

### Generate pie chart grid for contract report 

1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
plot_grid(piechart_reads, piechart_species_plot,
          ncol=2,
          align = "vh"
          )

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Category_breakdown.png", width=14, height=4)
```


## Calculate and plot relative abundance by category

1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
### Calculate the reads per category per sample
raw_reads %>%
  
  ## group by sample to calculate total reads per sample
  dplyr::group_by(sampleID) %>%
  mutate(total = sum(reads)) %>%
  
  ## group by sample and category to generate relative abundance of that category in each sample
  dplyr::group_by(sampleID, Category) %>%
  reframe(group_sum = sum(reads),
         group_relab = group_sum/total
           ) %>% distinct() %>%
  
### plot those values 
ggplot(., aes(y=group_relab, x=Category)) + 
  geom_boxplot(outlier.shape = NA, aes(color=Category), fill=NA) +
  geom_jitter(aes(fill=Category), width=0.2, shape=21, color='black', size = 0.75, alpha=0.35) +
    scale_color_manual(values = fill_colors) +
    scale_fill_manual(values = fill_colors) +
    labs(color = "Category") +
    scale_x_discrete(labels = scales::label_wrap(10), limits = categories) +
    theme_bw() + 
    xlab("Category") + ylab("Relative abundance") +
    theme(panel.background=element_rect(fill='white', colour='black'),
          legend.position = "none",
        axis.text.y = element_text(size=7, color="grey30"),
        axis.text.x = element_text(size=7), # angle=45, hjust=1
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=11, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=11, face="bold"))

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Categories_relative_abundance.png", width = 6.5, height = 4)
```


## Caluclate Top 30 Species List and plot  

1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
### Create dataframe that calculates the total reads per species and log10 transforms that value 
top_list <- raw_reads %>%
  filter(!Category == "Other" & !Category == "Livestock" & !Category == "Unassigned" & !Category == "Human") %>%
  group_by(Species_name, Common_name) %>%
  summarise(total = sum(reads),
            log = log10(total)) %>% 
  arrange(desc(total)) %>%
  head(30) %>% dplyr::select(-total)

ggplot(top_list, aes(x = fct_reorder(Common_name, log), y = log)) +
  geom_segment(aes(xend = Common_name, yend = 0), color = "#97C1DE") +  # Lollipop stick
  geom_point(size = 3, shape=21, color='grey30', fill = "#97C1DE") +  # Lollipop head
  coord_flip() +  # Flip coordinates for horizontal lollipop chart
  labs(
       x = "",
       y = "Normalized Reads") +
  
  theme_bw() +
  theme(axis.text.y = element_text(size = 8, color='black'), #, face="italic"
        axis.text.x = element_text(size = 6),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=10, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold"),
        axis.text.x.top = element_text(size = 8, color='black', face="italic", angle = 45, hjust = 0)) +
  scale_y_continuous(
    labels = comma,
    limits = c(0, max(top_list$log) + (max(top_list$log)*0.1))  # Set the upper limit to max value + 10%
  )

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Top_species_log.png", width=3.5, height=6)
```

## Bubble plot 

1. Change paths of output to desired folder (data/figures is the suggested data structure)  

```{r}
raw_reads %>%
  
  ## calculate the sum of reads in the million 
  ## add column 'xaxis' to plot everything on 1 tick mark on the x axis
  dplyr::group_by(Species_name, Common_name, Category) %>%
  reframe(sum = sum(reads)/1000000,
          xaxis = "x") %>%
  
  ## add taxonomy information
  left_join(., df_tax, by = "Species_name") %>%

  ## remove non-target categories
  filter(!Category == "Other" & !Category == "Livestock" & !Category == "Unassigned" & !Category == "Human") %>%

  # Create a factor for Common_name ordered by Order within each Category; then ordered by PhylogenySorting
  dplyr::group_by(Category) %>%
  mutate(Common_name = factor(Common_name, levels = unique(Common_name[order(Order, desc(PhylogenySorting))]))) %>%
  ungroup() %>%

  ggplot(., aes(x=xaxis, y=Common_name)) + 

  geom_point(aes(size=sum, fill=sum), color = 'black', shape=21) +
  scale_fill_gradient(na.value = "white", low = "lightskyblue2", high = "#0C4D66") +

  facet_grid2(Category ~ ., scales = "free", space = "free") +

  theme_bw() +
  labs(
    x="",
    y="",
    fill = "Reads (M)",
    size = "Reads (M)"
  ) +

  theme(
    axis.text.y = element_text(size = 8, color = 'black'),
    axis.text.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=12, face="bold"),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=12, face="bold"),
    ## facet wrap labels
    strip.text.x = element_text(color = "black", face = "bold", size = 12),
    #strip.text.y = element_text(color = "black", face = "bold", size = 12, angle=0),
    strip.text.y = element_blank(),
    strip.background.y = element_blank(),
    strip.clip = "off",
    # Combine legends
    legend.position = "right",
    legend.box = "vertical"
  ) +
  
  guides(
    fill = "none",
    size = guide_legend(order = 2, reverse = TRUE,
                        override.aes = list(fill = scales::seq_gradient_pal("#0C4D66", "lightskyblue2")
                                            (seq(0, 1, length.out = 4))))
  )

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Species_bubbleplot.png", width=4.5, height=10)
```

## Biodiversity metrics 

Creating phyloseq object with the df_relative matrix.   
User edits: Change to df_raw if needed.  
  
```{r}
df_for_biodiv <- rel_ab_df %>% filter(!is.na(reads))

otu <- otu_table(df_for_biodiv %>% spread(sampleID, reads) %>%
                   subset(!Category == "Unassigned") %>%
                   dplyr::select(-Common_name, -Category) %>%
                   column_to_rownames(var = "Species_name"), 
                  taxa_are_rows = T)

# meta_phyloseq <- sample_data(
#   meta %>% ## rownames are also needed in phyloseq meta table
#   mutate(sampleID2=sampleID) %>% column_to_rownames(var = "sampleID2")
# )

## Merge metadata and OTU table into one phyloseq "object"
phylo_obj <- merge_phyloseq(otu)

## view phyloseq obj 
## expected output = otu_table() with taxa and sample numbers and sample_data() with the sample and column numbers
# print(phylo_obj)

# Ensure that your OTU table doesn't contain any NA or negative values (output should be FALSE)
any(is.na(otu_table(phylo_obj)))
any(otu_table(phylo_obj) < 0)
```

### Calculating alpha diversity and species richness 

This is currently done with relative abundance values which is not the best way.. But might be better than raw reads.

```{r}
alpha_div <- estimate_richness(phylo_obj, measures = c("Shannon", "Simpson")) %>%
  rownames_to_column(var = "sampleID")# %>% left_join(., meta, by = "sampleID") 

## Species Richness
biodiv_df <- rel_ab_df %>% subset(!Category == "Unassigned") %>% dplyr::group_by(sampleID) %>%
  filter(reads > 0) %>%
  reframe(Richness = n_distinct(Species_name)) %>%
  
  ## combining this information with alphadiversity 
  left_join(., alpha_div, by = "sampleID") %>%
  
  mutate(across(3:4, ~ round(.x, 3)))

biodiv_df %>% 
  write_xlsx("docs/eDNA 12S metab/example_output/Biodiversity.xlsx")
```

Plotting

```{r}
## plotting
biodiv_df %>%
  ggplot(., aes(x=Richness, y=Shannon)) + 
  
  annotate("rect", xmin=quantile(biodiv_df$Richness, 0.75), xmax=Inf,
             ymin=quantile(biodiv_df$Shannon, 0.75, na.rm = TRUE), ymax=Inf,
             alpha=0.2, fill="#058C42") +

  geom_point(fill = "#97C1DE", color='black', shape=21, alpha=0.5, size=2) + 
  
  labs(
    x = "Sample Richness",
    y = "Sample Evenness (Shannon Index)"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size=8, color="grey30"),
        axis.text.x = element_text(size=8, color="grey30"),
        axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=10, face="bold"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold")
  )

ggsave("docs/eDNA 12S metab/example_output/figures/SampleReport_Biodiversity.png", width = 5.5, height = 4)
```





